---
title: "Chapter 2: Vectors, Part 1"
output: html_document
date: "`r Sys.Date()`"
---

## Code Exercises

Exercise 2-1.

```{r}
V <- c(1,2)
W <- c(4,-6)
VW <- V + W
plot(NA, xlim = c(-1,6), ylim = c(-6,3))
arrows(0,0,V[1],V[2], length = 0.2, col = "blue")
arrows(V[1],V[2],VW[1], VW[2], length = 0.2, col = "red")
arrows(0,0,VW[1],VW[2], length = 0.2, col = "green")
```

```{r}
V <- c(1,2)
W <- c(4,-6)
VW <- V - W
plot(NA, xlim = c(-1,6), ylim = c(-6,3))
arrows(0,0,V[1],V[2], length = 0.2, col = "blue")
arrows(0,0,W[1], W[2], length = 0.2, col = "red")
arrows(W[1],W[2],V[1],V[2], length = 0.2, col = "green")
```

Exercise 2-2.

Norm of vectors

```{r}
V <- c(1,2,3,7,8,9)
vnorm <- V^2 |> sum() |> sqrt()
vnorm
```

```{python}
import numpy as np
v = np.array([1,2,3,7,8,9])
v_dim = len(v) # math dimensionality
v_mag = np.linalg.norm(v) # math magnitude, length, or norm
v_mag
```

Exercise 2-3.

```{r}
V <- c(-7,6)
vnorm <- V^2 |> sum() |> sqrt()
vunit <- V / vnorm
vunit
```

Exercise 2-4.

```{r}
normv <- function(v) {
    return( v^2 |> sum() |> sqrt() )
}

unitv <- function(v) {
    return( v / normv(v) )
}

unitMag <- function(v,m) {
    return( m * unitv(v) )
}

V <- c(1,2)
unitMag(V,4)
```

Exercise 2-6.

you can compute the squared norm of a vector as the dot product of that vector with itself. Look back to Equation 2-8 to convince yourself of this equivalence. Then confirm it using code.

```{r}
v <- c(1,2,3,7,8,9)

normv(v) ^ 2

v %*% v
```

Exercise 2-7.

Write code to demonstrate that the dot product is commutative. Commutative means that a × b = b × a, which, for the vector dot product, means that aTb = bTa. After demonstrating this in code, use equation Equation 2-9 to understand why the dot product is commutative.

```{r}
a <- c(1,-2,3,-4)
b <- c(-5,6,-7,8)
a %*% b
b %*% a
```

Exercise 2-9.

Implement orthogonal vector decomposition. Start with two random-number vectors t and r, and reproduce Figure 2-8 (note that your plot will look somewhat different due to random numbers). Next, confirm that the two components sum to t and that tTr and t\|\|r are orthogonal.

```{r}
r <- c(1.2,-0.05)
t <- c(0.8, -0.8)

paradecomp <- function(r,t) {
    return( r %*% ((t(t) %*% r) / (t(r) %*% r)) )
}

perpendecomp <- function(r,t) {
    return( t - paradecomp(r,t) )
}

tpara <- paradecomp(r,t)
tper <- perpendecomp(r,t)

# plots

plot(NA, xlim = c(-0.1,1.3), ylim = c(-1,0.2))
segments(x0 = 0, y0 = 0, x1 = r[1], y1 = r[2], col = "blue")
segments(x0 = 0, y0 = 0, x1 = t[1], y1 = t[2], col = "blue")
segments(x0 = 0, y0 = 0, x1 = tpara[1], y1 = tpara[2], col = "red", lwd = 5)
segments(x0 = 0, y0 = 0, x1 = tper[1], y1 = tper[2], col = "green", lwd = 5)
segments(x0 = t[1], y0 = t[2], x1 = tpara[1], y1 = tpara[2], col = "black", lwd = 5)
```

```{r}
library(ggplot2)
```

```{r}
ggplot() +
    geom_segment(aes(x = 0, y = 0, xend = r[1], yend = r[2]), 
                 color = "grey", size = 2) +
    geom_segment(aes(x = 0, y = 0, xend = t[1], yend = t[2]), size = 2) +
    geom_segment(aes(x = 0, y = 0, xend = tpara[1], yend = tpara[2]), 
                 linetype="dashed", size = 2) +
    geom_segment(aes(x = 0, y = 0, xend = tper[1], yend = tper[2]), 
                 linetype="dotted", size = 2) +
    # geom_segment(aes(x = t[1], y = t[2], xend = tpara[1], yend = tpara[2]), a = 0.5) +
    theme_classic()
```

