---
title: "Chapter 3: Vectors, Part 2"
output: html_document
date: "`r Sys.Date()`"
---

Linear Weighted Combination

```{r}
l1 <- 1
l2 <- 2
l3 <- -3
v1 <- c(4,5,1)
v2 <- c(-4,0,-4)
v3 <- c(1,3,2)
l1*v1 + l2*v2 + l3*v3
```

```{python}
import numpy as np
l1 = 1
l2 = 2
l3 = -3
v1 = np.array([4,5,1])
v2 = np.array([-4,0,-4])
v3 = np.array([1,3,2])
l1*v1 + l2*v2 + l3*v3
```


## Code Exercises

Exercise 3-1.

Rewrite the code for linear weighted combination, but put the scalars in a list and the
vectors as elements in a list (thus, you will have two lists, one of scalars and one of
NumPy arrays). Then use a for loop to implement the linear weighted combination
operation. Initialize the output vector using np.zeros(). Confirm that you get the
same result as in the previous code.

```{r}
v1 <- c(4,5,1)
v2 <- c(-4,0,-4)
v3 <- c(1,3,2)

weights <- c(1,2,-3)

vecs <- matrix(c(v1,v2,v3), nrow = 3)

weights
vecs

lwc <- matrix(0,1,3)

for (i in 1:ncol(vecs)) {
    lwc <- lwc + (weights[i] * vecs[,i])
}
lwc
```

```{python}
import numpy as np
weights = [1,2,-3]

v1 = np.array([4,5,1])
v2 = np.array([-4,0,-4])
v3 = np.array([1,3,2])

vecs = np.array([v1,v2,v3]).T

lwc = np.zeros(3)

for i in range(vecs.shape[1]):
    lwc += weights[i] * vecs[:,i]

lwc
```

Exercise 3-3.

```{r}
vec <- c(1,3)
n <- 100
scalars <- runif(n = n, min = -4, max = 4)
scalars

m <- matrix(scalars[1] * vec, nrow = 1)

for (i in 2:nrow(scalars)) {
    m <- rbind(m, scalars[i] * vec)
}
```

```{r}
plot_ly(x = m[,1], y = m[,2])
```

repeat the procedure but using two vectors in R3.

```{r}
v1 <- c(3,5,1)
v2 <- c(0,2,2)

scalars <- matrix(data = runif(200, min = -4, max = 4), ncol = 2)

p1 <- matrix(scalars[1,1] * v1, nrow = 1)
p2 <- matrix(scalars[1,2] * v2, nrow = 1)

for (i in 2:nrow(scalars)) {
    p1 <- rbind(p1, scalars[i,1] * v1)
    p2 <- rbind(p2, scalars[i,2] * v2)
}

p3 <- rbind(p1, p2)
```

```{r}
plot_ly(x = p3[,1], y = p3[,2], z = p3[,3], type="scatter3d", mode = "markers")
```

repeat the R3 case but setting the second vector to be 1/2 times the first.

```{r}
v1 <- c(3,5,1)
v2 <- 0.5 * v1

scalars <- matrix(data = runif(200, min = -4, max = 4), ncol = 2)

p1 <- matrix(scalars[1,1] * v1, nrow = 1)
p2 <- matrix(scalars[1,2] * v2, nrow = 1)

for (i in 2:nrow(scalars)) {
    p1 <- rbind(p1, scalars[i,1] * v1)
    p2 <- rbind(p2, scalars[i,2] * v2)
}

p3 <- rbind(p1, p2)
```

```{r}
plot_ly(x = p3[,1], y = p3[,2], z = p3[,3], type="scatter3d", mode = "markers")
```