---
title: "Chapter 10: Row Reduction and LU Decomposition"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
library(Matrix)
library(pracma)
```


## Working with Matrix Equations

### R

```{r}
A <- matrix(data = rnorm(4*4), nrow = 4)
B <- matrix(data = rnorm(4*4), nrow = 4)
```

```{r}
# solve for X
X1 <- solve(A) %*% B
X2 <- B %*% solve(A)
X1
X2
```

```{r}
# residual (should be zeros matrix)
res1 <- A %*% X1 - B
res2 <- A %*% X2 - B
res1 |> round(14)
res2 |> round(14)
```

### Python

```{python}
import numpy as np

A = np.random.randn(4,4)
B = np.random.randn(4,4)
```

```{python}
# solve for X
X1 = np.linalg.inv(A) @ B
X2 = B @ np.linalg.inv(A)
print(X1)
print(X2)
```

```{python}
# residual (should be zeros matrix)
res1 = A @ X1 - B
res2 = A @ X2 - B
print(res1.round(14))
print(res2.round(14))
```

## Row Reduction

"The goal of row reduction is to transform a dense matrix into an upper-triangle matrix."

## Gaussian Elimination

## Gauss-Jordan Elimination

### Python

```{python}
import sympy as sym
import numpy as np

# the matrix converted to sympy
M = np. array([[1,1,4], [-1/2,1,2]])
symMat = sym.Matrix(M)
```

```{python}
symMat.rref()[0]
```

### R

```{r}
M <- matrix(data = c(1, -1/2, 1, 1), nrow = 2)
b <- c(4, 2)
matlib::gaussianElimination(M, b)
```

## LU Decomposition

"LU" stands for "lower upper". The idea is to decompose a matrix into the product of two triangular matrices.

A = LU

```{r}
# numerical example

L <- matrix(data = c(1,1/2,1,0,1,1,0,0,1), nrow = 3)
U <- matrix(data = c(2,0,0,2,-1,0,4,1,3), nrow = 3)
print("Lower")
L
print("Upper")
U
L %*% U
```

## Row Swaps via Permutation Matrices

Row swaps are one of the tricks of row reduction, and are implemented through a permutation matrix.

```{r}
mat1 <- matrix(data = c(3,0,0,2,0,7,1,5,2), nrow = 3)
mat1

pmat <- matrix(data = c(1,0,0,0,0,1,0,1,0), nrow = 3)
pmat

pmat %*% mat1
```

```{r}
heatmap(mat1, Rowv = NA, Colv = NA, scale = "none")
heatmap(pmat, Rowv = NA, Colv = NA, scale = "none")
heatmap(pmat %*% mat1, Rowv = NA, Colv = NA, scale = "none")
```

Exercise 10-1.

```{r}
mat1 <- matrix(data = rnorm(100*100), nrow = 100)
```

```{r}
lumat <- Matrix::lu(mat1)
```

Exercise 10-2.

```{r}
mat1 <- matrix(data = rnorm(3*6), nrow = 6)
Matrix::rankMatrix(mat1)[1]
mat2 <- matrix(data = rnorm(3*8), nrow = 3)
Matrix::rankMatrix(mat2)[1]
mat3 <- mat1 %*% mat2
Matrix::rankMatrix(mat3)[1]
```

```{r}
luMat <- Matrix::lu(mat3)
L <- Matrix::expand(luMat)$L |> round(14) |> as.matrix()
U <- Matrix::expand(luMat)$U |> round(14) |> as.matrix()
```

```{r}
heatmap(mat3, Rowv = NA, Colv = NA, scale = "none")
heatmap(L, Rowv = NA, Colv = NA, scale = "none")
heatmap(U, Rowv = NA, Colv = NA, scale = "none")
```

Exercise 10-3.

```{r}
mat1 <- matrix(data = rnorm(5*5), nrow = 5)
lumat <- Matrix::lu(mat1)
U <- Matrix::expand(lumat)$U |> as.matrix()
```


```{r}
diag(U) |> prod()
Matrix::det(mat1)
```

Exercise 10-4.

```{r}
n = 4
A <- matrix(data = rnorm(n*n), nrow = n)
A
```

```{r}
lumat <- Matrix::lu(A) |> Matrix::expand()
U <- as.matrix(lumat$U)
L <- as.matrix(lumat$L)
P <- as.matrix(lumat$P)
```

```{r}
inv(U) %*% inv(L) %*% t(P)
inv(A)
```

```{r}
A %*% inv(A) |> round(14)
```

```{python}
import numpy as np

A = np.random.rand(4,4)
A
```

```{python}
import scipy

lumat = scipy.linalg.lu(A)
P = lumat[0]
L = lumat[1]
U = lumat[2]
```

```{python}
np.linalg.inv(U) @ np.linalg.inv(L) @ P.T
np.linalg.inv(A)
```

```{python}
mat = A @ np.linalg.inv(A)
mat.round(14)
```

Exercise 10-5.

```{r}
n = 4
A <- matrix(data = rnorm(n*n), nrow = n)
A
```

```{r}
lumat <- Matrix::lu(A) |> Matrix::expand()
U <- as.matrix(lumat$U)
L <- as.matrix(lumat$L)
P <- as.matrix(lumat$P)
```

```{r}
t(U) %*% t(L) %*% L %*% U
t(A) %*% A
```
