---
title: "Chapter 13: Eigendecomposition"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
```

An eigenvector means that matrix-vector multiplication acts like scalar-vector multiplication.

## Finding Eigenvalues/vectors

eigenvalues are like keys that you insert into the matrix to unlock the mystical eigenvector.

```{r}
mat1 <- matrix(data = c(1,3,2,4), nrow = 2)
eigen(mat1)
```

```{python}
import numpy as np

mat1 = np.array([[1,2], [3,4]])

np.linalg.eig(mat1)[0]
np.linalg.eig(mat1)[1]
```

```{r}
matrix(c(1,2,2,1), nrow = 2) - diag(x = 3, 2)
matrix(c(-2,2,2,-2), nrow = 2) %*% c(1,1)
```

```{r}
matrix(c(1,2,2,1), nrow = 2) - diag(x = -1, 2)
matrix(c(-2,2,2,-2), nrow = 2) %*% c(1,1)
```


## The Special Awesomeness of Symmetric Matrices

### Orthogonal Eigenvectors

All eigenvectors of a symmetric matrix are pair-wise orthogonal.

```{r}
# random symmetric matrix
A <- matrix(data = sample(-3:4, 9, replace = T), nrow = 3)
A <- t(A) %*% A

# its eigendecomposisiton
L <- eigen(A)$values
V <- eigen(A)$vectors

# all pairwise dot products
V[,1] %*% V[,2] |> round(15)
V[,1] %*% V[,3] |> round(15)
V[,2] %*% V[,3] |> round(15)

```

### Real-Valued Eigenvalues

A second special property of symmetric matrices is that they have real-valued eigenvalues (and therefore real-valued eigenvectors).

```{r}
A <- matrix(data = c(-3,3,0,-3,-2,1,0,3,2), nrow = 3)
L <- eigen(A)$values
V <- eigen(A)$vectors

L
```

```{r}
A <- matrix(data = c(-3,-3,0,-3,-2,1,0,1,2), nrow = 3)
L <- eigen(A)$values
V <- eigen(A)$vectors

L
```

## Eigendecomposition of Singular Matrices

```{r}
A <- matrix(data = 1:9, nrow = 3)
L <- eigen(A)$values
V <- eigen(A)$vectors

pracma::Rank(A)
L |> round(2)
V |> round(2)
```

## Generalized Eigendecomposition (GED)

```{r}
A <- matrix(data = rnorm(16), nrow = 4)
A <- A %*% t(A)

B <- matrix(data = rnorm(16), nrow = 4)
B <- B %*% t(B) + A/10
A
B

C <- A %*% solve(B)

# GED
eigen(A, symmetric = T)$vectors
eigen(B, symmetric = T)$ vectors

geigen::geigen(A,B)
```

## Code Exercises

### Exercise 13-1.

```{r}
A <- matrix(data = rnorm(5*5), nrow = 5)
Ainv <- solve(A)

# eigenvectors of A == A inverse
eigen(A)$vectors
eigen(Ainv)$vectors[,5:1]

# eigenvalues of A == A inverse^-1
eigen(A)$values
eigen(Ainv)$values^-1 |> rev()

# eigenvalues of A^-1 == A inverse
eigen(A)$values^-1
eigen(Ainv)$values |> rev()
```

### Exercise 13-2.

```{r}
V1 <- c(-0.9,-0.4)
V2 <- c(-0.4, -0.9)

```

```{r}
vmat <- cbind(V1,V2)
egn <- eigen(vmat)
egn

V1 %*% egn$vectors
```


```{r}
plot(x = c(-1.5,1.5), y = c(-1.5,1.5), type="n", xlab="X", ylab="Y")
abline(v=0, h=0, col="gray")
matlib::vectors(rbind(V1,V2), col=c("black", "red"))
```

### Exercise 13-3.

```{r}
# additive method for symm matrix
addSym <- function(mat) {
    output <- (mat + t(mat)) / 2
}
```

```{r}
swapLargest2 <- function(vec) {
    temp <- unique(vec)
    temp <- sort(temp, decreasing = T)[1:2]
    first <- which(vec == temp[1])
    second <- which(vec == temp[2])
    vec[first] <- temp[2]
    vec[second] <- temp[1]
    return(vec)
}

swapSmallest2 <- function(vec) {
    temp <- unique(vec)
    temp <- sort(temp, decreasing = F)[1:2]
    first <- which(vec == temp[1])
    second <- which(vec == temp[2])
    vec[first] <- temp[2]
    vec[second] <- temp[1]
    return(vec)
}
```

```{r}
temp <- sample(-10:10, 5, replace = T)
temp
swapLargest2(temp)
swapSmallest2(temp)
```

```{r}
A <- matrix(data = rnorm(5*5), nrow = 5)
A <- addSym(A)
A
Aeval <- eigen(A)$values
reval <- sample(Aeval, 5)
reval
```

```{r}
# reconstruct the original matrix as VAV^-1

eigdag <- diag(eigen(A)$values)
eigvec <- eigen(A)$vectors

mat0 <- eigvec %*% eigdag %*% solve(eigvec) 
A
(A - mat0) |> norm(type = "F")
```

```{r}
# attempt to reconstruct the matrix using shuffled eigen values
eigdag <- diag(reval)
eigvec <- eigen(A)$vectors

mat1 <- eigvec %*% eigdag %*% solve(eigvec)
mat1
A
(A - mat1) |> norm(type = "F")
```

```{r}
# What happens if you only swap the two largest eigenvalues instead of randomly reordering them?
reval2 <- swapLargest2(Aeval)
eigdag <- diag(reval2)
eigvec <- eigen(A)$vectors

mat2 <- eigvec %*% eigdag %*% solve(eigvec) 
mat2
A
(A - mat2) |> norm(type = "F")
```

```{r}
# What happens if you only swap the two smallest eigenvalues instead of randomly reordering them?
reval3 <- swapSmallest2(Aeval)
eigdag <- diag(reval3)
eigvec <- eigen(A)$vectors

mat3 <- eigvec %*% eigdag %*% solve(eigvec) 
mat3
A
(A - mat3) |> norm(type = "F")
```

```{r}
dists <- c((A - mat0) |> norm(type = "F"),(A - mat1) |> norm(type = "F"),(A - mat2) |> norm(type = "F"),(A - mat3) |> norm(type = "F"))

types <- c("None", "All", "Largest two", "Smallest two")

df <- data.frame(types, dists)
```

```{r}
df |>
    mutate(types = factor(types, levels = c("None", "All", "Largest two", "Smallest two"))) |>
    ggplot(aes(x = types, y = dists)) +
    geom_col()
```

### Exercise 13-4.

compute 123 random 42 × 42 matrices, extract their eigenvalues, 
divide by the square root of the matrix size (42), plot the eigenvalues on the
complex plane.

#### Using torch

```{r}
library(torch)

x <- torch_randn(c(42,42,123))
```

```{r}
eigvals <- c()
index <- 1
for (i in 1:dim(x)[3]) {
    vals <- eigen(x[,,i])$values
    valsdiv <- vals / sqrt(42)
    eigvals <- append(eigvals, valsdiv)
}
```

```{r}
plot(eigvals, asp = 1)
```

#### Using base R

```{r}
dims <- c(42,42,123)
x <- array(data = rnorm(42*42*123), dim = dims)

eigvals <- c()
index <- 1
for (i in 1:dim(x)[3]) {
    vals <- eigen(x[,,i])$values
    valsdiv <- vals / sqrt(42)
    eigvals <- append(eigvals, valsdiv)
}

plot(eigvals, asp = 1)
```

### Exercise 13-5.

```{r}
x <- matrix(data = rnorm(9), nrow = 3)
symx <- t(x) %*% x
eigx <- eigen(symx)

for (i in 1:3) {
    nullV <- svd(symx - eigx$values[i] * diag(3))$d
    r <- cor(nullV, eigx$vectors[i,])
    sprintf("Correlation between N(A-lI) and evec %d: %0.2f", i, abs(r)) |> print()
}

```

```{python}
import numpy as np
from scipy.linalg import null_space

# Create a symmetric matrix
N = 3
A = np.random.randn(N,N)
A = A @ A.T

# eigendecompose
evals,evecs = np.linalg.eig(A)

# compare the eigenvectors with N(A-lI)
for i in range(N):

  # get the null space vector of the shifted matrix
  nullV = null_space( A-evals[i]*np.eye(N) )

  # check for a match with the eigenvector via correlation (normalizes for magnitudes)
  r = np.corrcoef(nullV.T,evecs[[i],:])[0,1]

  # and print (abs(r))
  print(f'Correlation between N(A-lI) and evec {i}: {np.abs(r):.2f}')

```

### Exercise 13-6.

```{r}
randmat <- diag(1:4)

qmat <- matrix(data = rnorm(4*4), nrow = 4) |> qr() |> qr.Q()

qmat %*% randmat %*% t(qmat)
```

### Exercise 13-7.

```{r}
dat1 <- read.csv("SeoulBikeData.csv", fileEncoding="latin1")
dat1$Date <- lubridate::dmy(dat1$Date)
```

```{r}
X <- dat1 |>
    select(Rainfall.mm., Temperature..C.) |>
    mutate(intercept = 1,
           v4 = (0.4 * Temperature..C.) + (4 * Rainfall.mm.)) |>
    as.matrix()

y <- dat1$Rented.Bike.Count

id <- diag(ncol(X))

gamma1 <- 0
gamma2 <- 0.01
```

```{r}
solve(crossprod(X) + gamma1 * id, tol = 1e-18) |> pracma::Rank()
solve(crossprod(X) + gamma2 * id, tol = 1e-18) |> pracma::Rank()
```

```{r}
beta <- solve(crossprod(X) + gamma1 * mean(eigen(t(X) %*% X)$values) * id, tol = 1e-18) %*% t(X) %*% y
pred <- X %*% beta
ssr <- sum((y - pred)^2)
sst <- sum((y - mean(pred))^2)
R2 <- ssr / sst
R2
```

```{r}
gammas <- seq(0, 0.2, 0.01)
r2s <- rep(NA, length(gammas))

for (i in 1:length(gammas)) {
    beta <- solve(crossprod(X) + gammas[i] * mean(eigen(t(X) %*% X)$values) * id, tol = 1e-18) %*% t(X) %*% y
    pred <- X %*% beta
    ssr <- sum((y - pred)^2)
    sst <- sum((y - mean(pred))^2)
    r2s[i] <- ssr / sst
}
```

```{r}
r2s |> plot()
```

### Exercise 13-8.

```{r}
X <- matrix(data = rnorm(3*10000), nrow = 3)
cormat <- matrix(data = c(1,.2,.9,.2,1,.3,.9,.3,1), ncol = 3)

eigcor <- eigen(cormat)
VA <- eigcor$vectors %*% sqrt(diag(eigcor$values))

Y <- VA %*% X

cormat2 <- cor(t(Y))
cormat2 |> round(1)
cormat
```

### Exercise 13-9.

```{r}
t(Y) %*% VA |> cor() |> round(1)
```

### Exercise 13-10

```{r}
n = 5
mat1 <- matrix(data = rnorm(n^2), nrow = n)
mat2 <- matrix(data = rnorm(n^2), nrow = n)
A <- t(mat1) %*% mat1
B <- t(mat2) %*% mat2
eigvecs <- geigen::geigen(A, B)$vectors
```

```{r}
# eigenvectors times their transposes
VV <- t(eigvecs) %*% eigvecs
VV

VBV <- t(eigvecs) %*% B %*% eigvecs
VBV |> round(14)
```

```{r}
heatmap(VV, Rowv = NA, Colv = NA, scale = "none")
heatmap(VBV, Rowv = NA, Colv = NA, scale = "none")
```

### Exercise 13-11.

```{r}
mat1 <- matrix(data = sample(-14:14, 4*4, replace = T), nrow = 4)
mat1
```

```{r}
eigmat <- eigen(mat1)
V <- eigmat$vectors * pi
D <- diag(eigmat$values)
Vi <- solve(eigmat$vectors)

A <- eigmat$vectors %*% D %*% Vi
Re(A)
mat1
```

